### 2017-11-13 vipassana-for-hackers.tex
fixing inline references from quote to italics

A 10-day Vipassana course requires that all students take the standard ``5 Precepts'' \cite{fiveprecepts}, as per Theravada. I do not recommend getting drunk, raping and murdering your fellow students, and stealing their belongings. However, the four precepts which forbid those behaviours are quite clearly external, making them components of course administration and therefore not applicable to this document.

@@

The hacker community has been on the verge of this exploration for quite some time. The easiest and most recent examples of this are no doubt the works of Douglas Hofstadter. ``Gödel, Escher, Bach: An Eternal Golden Braid'' captured the imaginations of the hacker community and grew from esoteric pub conversation piece to dinner party staple, around the same time as the term ``blogging'' evolved from portmanteau to an accessible part-time career for stay-at-home parents. The term ``strange loop'', coined in this book, has since become a mild obsession of philosophical and artistic computer scientists, linguists, and mathematicians, even inspiring the namesake software conference held in St. Louis. Hofstadter's more cohesive ``I Am a Strange Loop'' tackles the problem of consciousness more succinctly. However, other than directing the reader to Zen\footnote{Zen is relevant because the practice of Zazen (Zen meditation) is a nervous-system-based meditation similar to Vipassana. The reason I do not discuss Zazen in detail or compare it to Vipassana in this paper is because Zazen, in my experience, is not systematic. The subject of inspection — the nervous system — remains, but Zen teachers tend to avoid explicit instruction, leaving the student to explore on her own. The outcomes, therefore, appear similar but watered down.} at multiple points, these books do not invite the reader to explore these concepts directly. There is no hacking. A thorough read of GEB or IAaSL is akin to reading all four volumes of Knuth's ``The Art of Computer Programming'' without having ever touched a computer or to reading ``The Joy of Cooking'' cover to cover without having ever set foot inside a kitchen.

@@

\begin{itemize}
\item Homoiconicity: the language is represented in itself and code written in the language is understood by the language
\item Metaprogramming: the language can program itself and understand itself
\item Self-mutation and self-replication: a LISP program can inherently alter its own internal behaviour and easily write entirely new programs
\item Recursion and trampolining: mathematical functional recursion and mutual recursion between two functions are natural operations in LISP
\end{itemize}

@@

Vipassana would appear to be ``turtles all the way down'' in precisely this sense. Vipassana is the process of digging through your consciousness using consciousness itself. At no point do you reach the machine beneath the machine: ``Aha! I found it! The Consciousness Machine. Now I can do away with the meditation and just read the manufacturer's open specification and I'm sorted.'' The machine at the bottom of your consciousness \textbf{is} a strange loop. Having a look at this strange loop is a lot more fun than reading about it.

@@

Abrash discusses the nature of the ``Matrix Paradox'' (more commonly known as the ``Simulation Hypothesis'' \cite{simulationhypothesis}) as simply the right neurons firing to simulate any particular environment. ``How do we know we are not in the matrix?'' The question is valid, but the diagram is missing a great deal of sensory input. Figure~\ref{fig:michael-abrash-senses-extended} is a slightly better model of reality.

@@

The map for an anechoic chamber may seem surprising. However, just because the room you are in has a negative decibel rating does not mean you cannot hear. In fact, in this environment, your sense of hearing is likely to be amplified — and turned inward. You will hear your heart beating, your blood pumping, your digestive system sloshing around. The better the anechoic chamber at swallowing sound, the more intense this experience is going to be.

For this reason, the above anechoic chamber has a record of 45 minutes for the longest period anyone has sat inside it. As it turns out, hearing one's own bodily functions at deafening internal volumes is likely to cause someone to go a bit crazy, hence the highlighted bar for thought activity.

@@

I would describe the positive effects of nervous system meditations to live on a spectrum. On one end of the spectrum are the positive, rationality-enhancing qualities of calming one's nervous system with quiet attention. On the other end of the spectrum are the subtler qualities: Meditation seems to make a person more generous, more compassionate, more empathetic. Why would systematic analysis of the autonomic nervous system have these effects? A wide body of research suggests that many forms of meditation, including Vipassana, exhibit these positive effects. Many often identify specifically enhanced activity in the prefrontal cortex \cite{differentialengagement, meditationbrainstructure}, which is identified as the centre of the will to live and altruistic tendencies in the human brain. \cite{personalityneuroscience} However, literature explaining why or how such effects may occur is limited.

@@

If Vipassana is hacking the mind, is there an opportunity for hacking Vipassana? Can we hack the hacking process? Let's ask Martin Thompson. The following slide in Figure~\ref{fig:load-test} is from his 2013 talk \textit{Performance Testing Java Applications} \cite{testingjava} and while I have no intention of ostracizing non-computer hackers, software was my career once upon a time and the example not only appropriate but reasonably accessible.

@@

This particular distraction may feel justified as the reification of your skepticism. While it pays to be skeptical, try your hardest to avoid skepticism during the course. Approach the practice skeptically until you decide to take a course and then save your analysis of whatever newfound skepticism the course generates for you until after the course is completed. There will be plenty of time to discuss the meditation and the outcomes of the meditation on the 10th day when talking is permitted; explore your skepticism here. Other meditators are likely to be the only ones who will quite comprehend the concepts you are trying to discuss.

@@

@@

>>>>

A 10-day Vipassana course requires that all students take the standard \textit{5 Precepts} \cite{fiveprecepts}, as per Theravada. I do not recommend getting drunk, raping and murdering your fellow students, and stealing their belongings. However, the four precepts which forbid those behaviours are quite clearly external, making them components of course administration and therefore not applicable to this document.

@@

The hacker community has been on the verge of this exploration for quite some time. The easiest and most recent examples of this are no doubt the works of Douglas Hofstadter. \textit{Gödel, Escher, Bach: An Eternal Golden Braid} captured the imaginations of the hacker community and grew from esoteric pub conversation piece to dinner party staple, around the same time as the term ``blogging'' evolved from portmanteau to an accessible part-time career for stay-at-home parents. The term ``strange loop'', coined in this book, has since become a mild obsession of philosophical and artistic computer scientists, linguists, and mathematicians, even inspiring the namesake software conference held in St. Louis. Hofstadter's more cohesive \textit{I Am a Strange Loop} tackles the problem of consciousness more succinctly. However, other than directing the reader to Zen\footnote{Zen is relevant because the practice of Zazen (Zen meditation) is a nervous-system-based meditation similar to Vipassana. The reason I do not discuss Zazen in detail or compare it to Vipassana in this paper is because Zazen, in my experience, is not systematic. The subject of inspection — the nervous system — remains, but Zen teachers tend to avoid explicit instruction, leaving the student to explore on her own. The outcomes, therefore, appear similar but watered down.} at multiple points, these books do not invite the reader to explore these concepts directly. There is no hacking. A thorough read of GEB or IAaSL is akin to reading all four volumes of Knuth's \textit{The Art of Computer Programming} \cite{taocp} without having ever touched a computer or to reading \textit{The Joy of Cooking} \cite{joyofcooking} cover to cover without having ever set foot inside a kitchen.

@@

\begin{itemize}
\item \textbf{Homoiconicity:} the language is represented in itself and code written in the language is understood by the language
\item \textbf{Metaprogramming:} the language can program itself and understand itself
\item \textbf{Self-mutation and self-replication:} a LISP program can inherently alter its own internal behaviour and easily write entirely new programs
\item \textbf{Recursion and trampolining:} mathematical functional recursion and mutual recursion between two functions are natural operations in LISP
\end{itemize}

@@

Vipassana would appear to be ``turtles all the way down'' in precisely this sense. Vipassana is the process of digging through your consciousness using consciousness itself. At no point do you reach the machine beneath the machine: ``Aha! I found it! The Consciousness Machine. Now I can do away with the meditation and just read the manufacturer's open specification and I'm sorted.'' \textit{The machine at the bottom of your consciousness is a strange loop}. Having a look at this strange loop is a lot more fun than reading about it.

@@

Abrash discusses the nature of the ``Matrix Paradox'' (more commonly known as the \textit{Simulation Hypothesis} \cite{simulationhypothesis}) as simply the right neurons firing to simulate any particular environment. ``How do we know we are not in the matrix?'' The question is valid, but the diagram is missing a great deal of sensory input. Figure~\ref{fig:michael-abrash-senses-extended} is a slightly better model of reality.

@@

The map for an anechoic chamber may seem surprising. However, just because the room you are in has a negative decibel rating does not mean you cannot hear. In fact, in this environment, your sense of hearing is likely to be amplified --- and turned inward. You will hear your heart beating, your blood pumping, your digestive system sloshing around. The better the anechoic chamber at swallowing sound, the more intense this experience is going to be.

For this reason, the above anechoic chamber at Orfield Laboratories \cite{orfield} has an unofficial record of 45 minutes for the longest period anyone has sat inside it. As it turns out, hearing one's own bodily functions at deafening internal volumes is likely to cause someone to go a bit crazy, hence the highlighted bar for thought activity.

@@

I would describe the positive effects of nervous system meditations to live on a spectrum. On one end of the spectrum are the positive, rationality-enhancing qualities of calming one's nervous system with quiet attention. On the other end of the spectrum are the subtler qualities: Meditation seems to make a person more generous, more compassionate, more empathetic. Why would systematic analysis of the autonomic nervous system have these effects? A wide body of research suggests that many forms of meditation, including Vipassana, exhibit these positive effects. Many often identify specifically enhanced activity in the prefrontal cortex \cite{differentialengagement, meditationbrainstructure}, which is identified as the centre of the will to live and altruistic tendencies in the human brain.\cite{personalityneuroscience} However, literature explaining why or how such effects may occur is limited.

@@

If Vipassana is hacking the mind, is there an opportunity for hacking Vipassana? Can we hack the hacking process? Let's ask Martin Thompson. The following slide in Figure~\ref{fig:load-test} is from his 2013 talk \textit{Performance Testing Java Applications} \cite{testingjava} and while I have no intention of ostracizing non-computer hackers, software was my career once upon a time and the example is not only appropriate but reasonably accessible.

@@

This particular distraction may feel justified as the reification of your skepticism. While it pays to be skeptical, try your hardest to avoid skepticism during the course. Approach the practice skeptically until you decide to take a course and then save your analysis of whatever newfound skepticism the course generates for you until after the course is completed. There will be plenty of time to discuss the meditation and the outcomes of the meditation on the tenth day when talking is permitted; explore your skepticism here. Other meditators are likely to be the only ones who will quite comprehend the concepts you are trying to discuss.

@@

\bibitem{taocp}
  Donald Knuth.
  \textit{The Art of Computer Programming}.
  Addison-Wesley, 1962, 1968, 1969, 1973, 2005, 2011, 2015, 2017.
  ISBN: 0-201-03801-3

\bibitem{joyofcooking}
  Irma S. Rombauer, Marion Rombauer Becker.
  \textit{The Joy of Cooking, Sixth Edition}.
  Bobbs-Merrill, 1975.
  ISBN: 0-02-604570-2

@@

\bibitem{orfield}
  Lindsey Galloway.
  \url{http://www.bbc.com/travel/story/20121022-the-quietest-place-on-earth}
  \textit{The quietest place on Earth}
  BBC, 23 October 2012.
